From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-07-22 dictionary.tex
Neron |--> N\'eron

\keywords{function-sheaf dictionary, trace of Frobenius, geometrization, categorification, character sheaves,smooth commutative group schemes,
finite fields, local fields, algebraic tori, Neron models, Greenberg functor, quasicharacters, quasicharacter sheaves.}

@@

Having defined quasicharacter sheaves on Neron models of commutative algebraic groups over $K$
and character sheaves on commutative group schemes over $\Fq$, we consider how
these categories are related as $K$ and $\Fq$ vary.  We describe (\S\ref{ssec:basechange}) functors between categories
of quasicharacter sheaves that model restriction and norm homomorphisms of character groups $G(k')^* \to G(k)^*$ and $G(k)^* \to G(k')^*$,
and describe how quasicharacter sheaves behave under Weil restriction (\S\ref{ssec:wrK}).  We also give (\S\ref{ssec:transfer})
a categorical version of a result of Chai and Yu \cite{chai-yu:01a},
relating quasicharacter sheaves for tori over different local fields, even local fields
with different characteristic.

@@

\subsection{N\'eron models}\label{ssec:neron}
We will consider connected commutative algebraic groups over $K$ that admit a N\'eron model,
by which we mean a locally finite type N\'eron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected
commutative algebraic groups over $K$ that contain no subgroup isomorphic to $\mathbb{G}_\text{a}$.
Write $\mathcal{N}_K$ for the full subcategory of the category of algebraic groups consisting of such objects.
This category is additive, and includes all algebraic tori over $K$, abelian varieties over $K$ and unipotent $K$-wound groups.
We write $\mathcal{N}$ for the category of Neron models that arise in this way; in particular, $\mathcal{N}$
is a full subcategory of the category of smooth commutative group schemes over $R$.

@@

On the other hand, if $K' = K(\sqrt{\pi})$ is totally ramified over $K$ then the N'eron model of $U_1(K'/K)$ is affine, namely
$\Spec{R[x,y] / (x^2 - \pi y^2 - 1)}$.  The special fiber is the the disjoint union of two affine lines.

Finally, we remark that if $K'/K$ is any finite extension of local fields and $X'$ is a N\'eron model for $X_{K'}$ then
$\Res_{R'/R}(X')$ is a N\'eron model for $\Res_{K'/K}(X_{K'}$ \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 7.6, Prop. 6}.
\end{example}

@@

 \begin{remark}
In Section~\ref{ssec:CS_tori} we see that the \'etale site on $\Gr_R(X)$ is rich enough to geometrize all
quasicharacters of $X(K)$ as $\ell$-adic local systems on $\Gr_R(X)$, where $X$ is a Neron model for an
algebraic torus or an abelian variety over a local field $K$. 
It is natural to ask if the \'etale site on the generic fibre $X_K$ would have sufficed.
This seems unlikely, since the geometric \'etale fundamental group of $\Gm{K}$ is ${\hat \ZZ}$;
however, limited results in this direction were established in \cite{Cunningham-Kamgarpour} when $K = \mathbb{Q}_p$.
\end{remark}

@@

As we explained in the Introduction, our original motivation for this paper was to find a
geometrization and categorification of quasicharacters of $p$-adic tori. 
This is now provided by the following adaptation of Theorem~\ref{thm:CSXK} in the case when
$T\in \mathcal{N}$ is a Neron model for an algebraic torus over $K$.

\begin{corollary}\label{cor:CS_tori}
Let $T$ be a Neron model for an algebraic torus over $K$.
The following is a commutative diagram of exact sequences.
\[
  \begin{tikzcd}[row sep=20, column sep=18]
{}  & 0 \arrow{d} & 0 \arrow{d} &  & \\ 
   & \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}  \arrow{d} & \arrow{d} \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}} & 0 \arrow{d} & \\
    0 \rar & \QCSiso{\pi_0(T)} \arrow{r}{\pi_0^*} \dar{\TrFrob{\pi_0(T)}}
    & \QCSiso{T} \arrow{r}{\iota_0^*}  \arrow{d}{\TrFrob{\Gr_R(T)}} & \QCSiso{T^0} \arrow{r} \arrow{d}{\TrFrob{\Gr_R(T)^0}} & 0\\
    0 \arrow{r} & \Hom(\pi_0(T)(\Fq), \EEx) \arrow{r}{\text{inf}} \arrow{d}{\iso}
    & \arrow{d}{\iso} \Hom_\text{}(T(K),\EEx) \arrow{r}{\text{res}} & \Hom_\text{}(T^0(R),\EEx) \arrow{r} \arrow{d}{\iso} & 0\\
 &  0  & 0 & 0 & 
  \end{tikzcd}
 \]
 \end{corollary}
\begin{proof} 
The horizontal sequence of groups coming from categories of quasicharacter sheaves is exact by
Proposition~\ref{prop:middleexact}, together with the observation that the functors $\pi_0^*$ and $\iota^*$ preserve limits.
It is elementary that the horizontal sequence of quasicharacters is exact.
Accordingly, by Theorem~\ref{thm:CSXK}, the kernel of the  is $\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{}}$. 
By \cite{bitan:discriminant}*{Eq 3.1}, the special fibre of the component group scheme for $T$ is given by
\[
 \pi_0(T)_{\bFq} = X_*(T)_{\mathcal{I}_K},
\]
where $X_*(T)$ is the cocharacter lattice of $T_K$ and $\mathcal{I}_K$ is the inertia group for $K$.
Thus,
\[
\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{\Fq}} = \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}.
\] 
Thus, the middle vertical sequence is exact.
Since $T^0$ and $\pi_0(T)$ do not lie in $\mathcal{N}$, we cannot use Theorem~\ref{thm:CSXK} to
determine the image and trace of Frobenius for these schemes. 
Instead, we observe that $T^0$ and $\pi_0(T)$ are smooth commutative group schemes over $R$,
so Definition~\ref{def:QCS} gives meaning to categories $\QCS(T^0)$ and $\QCS(\pi_0(T))$, and,
moreover, that $\pi_0(\Gr^R_n(T^0))= \pi_0(T^0)_{\Fq} = 1$ and $\pi_0(\Gr^R_n(\pi_0(T)) = \pi_0(T)_{\Fq}$ are both independent of $n$.
It follows that the vertical sequences through $\QCSiso{T^0}$ and $\QCSiso{\pi_0(T)}$ are exact by
Theorem~\ref{thm:snake} and Definition~\ref{def:QCS}.
The diagram commutes by Lemma~\ref{lem:pullback}.
\end{proof}

@@

\begin{corollary}
Let $T$ be a Neron model for an algebraic torus over $K$.
For $\cs{E}\in \QCS(\pi_0(T))$, $\cs{F}\in {\QCS(T)}$ and $\cs{F}^0\in {\QCS(T^0)}$, there are canonical isomorphisms
\[
\Aut(\cs{E})\iso (\check{T}_\ell)^{\Weil{K}},
\quad
\Aut(\cs{F})\iso (\check{T}_\ell)^{\Weil{K}},
\quad
\Aut(\cs{F}^0)\iso 1,
\]
 where $\check{T}_\ell \ceq \Hom(X_*(T),\EEx)$, the $\ell$-adic dual torus to $T_K$.
\end{corollary}

>>>>

\keywords{function-sheaf dictionary, trace of Frobenius, geometrization, categorification, character sheaves,smooth commutative group schemes,
finite fields, local fields, algebraic tori, N\'eron models, Greenberg functor, quasicharacters, quasicharacter sheaves.}

@@

Having defined quasicharacter sheaves on N\'eron models of commutative algebraic groups over $K$
and character sheaves on commutative group schemes over $\Fq$, we consider how
these categories are related as $K$ and $\Fq$ vary.  We describe (\S\ref{ssec:basechange}) functors between categories
of quasicharacter sheaves that model restriction and norm homomorphisms of character groups $G(k')^* \to G(k)^*$ and $G(k)^* \to G(k')^*$,
and describe how quasicharacter sheaves behave under Weil restriction (\S\ref{ssec:wrK}).  We also give (\S\ref{ssec:transfer})
a categorical version of a result of Chai and Yu \cite{chai-yu:01a},
relating quasicharacter sheaves for tori over different local fields, even local fields
with different characteristic.

@@

\subsection{N\'eron models}\label{ssec:neron}
We will consider connected commutative algebraic groups over $K$ that admit a N\'eron model,
by which we mean a locally finite type N\'eron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected
commutative algebraic groups over $K$ that contain no subgroup isomorphic to $\mathbb{G}_\text{a}$.
Write $\mathcal{N}_K$ for the full subcategory of the category of algebraic groups consisting of such objects.
This category is additive, and includes all algebraic tori over $K$, abelian varieties over $K$ and unipotent $K$-wound groups.
We write $\mathcal{N}$ for the category of N\'eron models that arise in this way; in particular, $\mathcal{N}$
is a full subcategory of the category of smooth commutative group schemes over $R$.

@@

On the other hand, if $K' = K(\sqrt{\pi})$ is totally ramified over $K$ then the N\'eron model of $U_1(K'/K)$ is affine, namely
$\Spec{R[x,y] / (x^2 - \pi y^2 - 1)}$.  The special fiber is the the disjoint union of two affine lines.

Finally, we remark that if $K'/K$ is any finite extension of local fields and $X'$ is a N\'eron model for $X_{K'}$ then
$\Res_{R'/R}(X')$ is a N\'eron model for $\Res_{K'/K}(X_{K'})$ \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 7.6, Prop. 6}.
\end{example}

@@

 \begin{remark}
In Section~\ref{ssec:CS_tori} we see that the \'etale site on $\Gr_R(X)$ is rich enough to geometrize all
quasicharacters of $X(K)$ as $\ell$-adic local systems on $\Gr_R(X)$, where $X$ is a N\'eron model for an
algebraic torus or an abelian variety over a local field $K$. 
It is natural to ask if the \'etale site on the generic fibre $X_K$ would have sufficed.
This seems unlikely, since the geometric \'etale fundamental group of $\Gm{K}$ is ${\hat \ZZ}$;
however, limited results in this direction were established in \cite{Cunningham-Kamgarpour} when $K = \mathbb{Q}_p$.
\end{remark}

@@

As we explained in the Introduction, our original motivation for this paper was to find a
geometrization and categorification of quasicharacters of $p$-adic tori. 
This is now provided by the following adaptation of Theorem~\ref{thm:CSXK} in the case when
$T\in \mathcal{N}$ is a N\'eron model for an algebraic torus over $K$.

\begin{corollary}\label{cor:CS_tori}
Let $T$ be a N\'eron model for an algebraic torus over $K$.
The following is a commutative diagram of exact sequences.
\[
  \begin{tikzcd}[row sep=20, column sep=18]
{}  & 0 \arrow{d} & 0 \arrow{d} &  & \\ 
   & \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}  \arrow{d} & \arrow{d} \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}} & 0 \arrow{d} & \\
    0 \rar & \QCSiso{\pi_0(T)} \arrow{r}{\pi_0^*} \dar{\TrFrob{\pi_0(T)}}
    & \QCSiso{T} \arrow{r}{\iota_0^*}  \arrow{d}{\TrFrob{\Gr_R(T)}} & \QCSiso{T^0} \arrow{r} \arrow{d}{\TrFrob{\Gr_R(T)^0}} & 0\\
    0 \arrow{r} & \Hom(\pi_0(T)(\Fq), \EEx) \arrow{r}{\text{inf}} \arrow{d}{\iso}
    & \arrow{d}{\iso} \Hom_\text{}(T(K),\EEx) \arrow{r}{\text{res}} & \Hom_\text{}(T^0(R),\EEx) \arrow{r} \arrow{d}{\iso} & 0\\
 &  0  & 0 & 0 & 
  \end{tikzcd}
 \]
 \end{corollary}
\begin{proof} 
The horizontal sequence of groups coming from categories of quasicharacter sheaves is exact by
Proposition~\ref{prop:middleexact}, together with the observation that the functors $\pi_0^*$ and $\iota^*$ preserve limits.
It is elementary that the horizontal sequence of quasicharacters is exact.
Accordingly, by Theorem~\ref{thm:CSXK}, the kernel of the  is $\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{}}$. 
By \cite{bitan:discriminant}*{Eq 3.1}, the special fibre of the component group scheme for $T$ is given by
\[
 \pi_0(T)_{\bFq} = X_*(T)_{\mathcal{I}_K},
\]
where $X_*(T)$ is the cocharacter lattice of $T_K$ and $\mathcal{I}_K$ is the inertia group for $K$.
Thus,
\[
\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{\Fq}} = \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}.
\] 
Thus, the middle vertical sequence is exact.
Since $T^0$ and $\pi_0(T)$ do not lie in $\mathcal{N}$, we cannot use Theorem~\ref{thm:CSXK} to
determine the image and trace of Frobenius for these schemes. 
Instead, we observe that $T^0$ and $\pi_0(T)$ are smooth commutative group schemes over $R$,
so Definition~\ref{def:QCS} gives meaning to categories $\QCS(T^0)$ and $\QCS(\pi_0(T))$, and,
moreover, that $\pi_0(\Gr^R_n(T^0))= \pi_0(T^0)_{\Fq} = 1$ and $\pi_0(\Gr^R_n(\pi_0(T)) = \pi_0(T)_{\Fq}$ are both independent of $n$.
It follows that the vertical sequences through $\QCSiso{T^0}$ and $\QCSiso{\pi_0(T)}$ are exact by
Theorem~\ref{thm:snake} and Definition~\ref{def:QCS}.
The diagram commutes by Lemma~\ref{lem:pullback}.
\end{proof}

@@

\begin{corollary}
Let $T$ be a N\'eron model for an algebraic torus over $K$.
For $\cs{E}\in \QCS(\pi_0(T))$, $\cs{F}\in {\QCS(T)}$ and $\cs{F}^0\in {\QCS(T^0)}$, there are canonical isomorphisms
\[
\Aut(\cs{E})\iso (\check{T}_\ell)^{\Weil{K}},
\quad
\Aut(\cs{F})\iso (\check{T}_\ell)^{\Weil{K}},
\quad
\Aut(\cs{F}^0)\iso 1,
\]
 where $\check{T}_\ell \ceq \Hom(X_*(T),\EEx)$, the $\ell$-adic dual torus to $T_K$.
\end{corollary}
